---
title: "STAT 107 Final Project: NBA Team Statistics and Winning"
author: "Team 24: Derek de Gracia, Shafin Kazi, Tiffany Huang, Tyler Wong"
date: "Due: 2025-12-05"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("00_requirements.R")
search()

```
Abstract:
Analytics and data-driven metrics have transformed modern NBA strategy, influencing roster decisions, offensive schemes, and defensive priorities. While many statistics are tracked in every game, not all contribute equally to winning. This study examines which team-level offensive and defensive statistics are most strongly associated with winning NBA games during the 2024–25 season.

Using cleaned box score data transformed into team aggregates, we evaluate the effect of shooting efficiency, three-point performance, free throws, defensive pressure (blocks + steals), and total points. We apply hypothesis testing, unsupervised clustering, and logistic regression to measure statistical significance, group team styles, and predict game outcomes. Results show that shooting efficiency (FG%) and defensive pressure are highly predictive of winning, while three-point accuracy contributes strongly but less consistently across teams. Logistic regression confirms that field-goal percentage and total points hold the highest predictive power with above-average classification accuracy.

This analysis provides insights useful to coaches, analysts, and scouting teams seeking to optimize roster construction and during-game strategy in the data-driven era of the NBA.

This report was made in conjunction with Derek De Gracia, Shafin Kazi, Tiffany Huang, Tyler Wong for STAT107 at UCR taught by Jose Sanchez Gomez. 
The repository is hosted by Github under https://github.com/shafinkazi/STAT107-NBA-Project

## 1. Introduction
The modern NBA is shaped heavily by analytics, efficiency, and data-driven decision-making. Teams increasingly rely on statistical insights to evaluate player performance, optimize game strategies, and gain competitive advantages. Among the many questions that arise in basketball analytics, one of the most fundamental is:

What statistical skills or attributes actually lead to winning games?

Although basketball is a team sport, game outcomes ultimately reflect patterns in offensive and defensive performance. Understanding which team-level statistics correlate most strongly with winning can benefit coaches, analysts, and front offices by highlighting which areas contribute most to success.

In this project, we analyze all team statistics from the 2024–25 NBA season to identify how offensive skills (such as 3-point shooting, assists, and field-goal efficiency) and defensive skills (such as steals, blocks, and turnovers) relate to game outcomes. By transforming player-level game logs into team-level summaries, we can compare performances across wins and losses and study which metrics are most predictive of success.

##Research Question
Which team-level offensive and defensive statistics are most strongly associated with winning NBA games in the 2024–25 season?

##To answer this question, we apply multiple statistical tools, including:
- Two-sample t-tests to compare offense vs. defense metrics between wins and losses
- K-means clustering to group team performance styles
- Logistic regression to predicti game outcomes from key features by turning the variables into binary conditions of win or lose

This multi-method approach allows us to evaluate individual statistical differences, build predictive models, and examine broader team-performance patterns.


## 2. Data
The data used for this analysis was sourced from Kaggle, a popular platform for data sharing that provides publicly available NBA teams' data in their playoffs. The dataset contains player-level box scores for NBA games during the 2024–25 season. Each row represents a single player's performance in one game and includes the offensive and defensive statistics used in our analysis:

Offensive metrics:
FG (field goals), FGA (field goal attempts), FG%, 3P (three-pointers made), 3PA (three-point attempts), 3P% (three-point percentage), AST (assists), and PTS (points).

Defensive metrics:
STL (steals), BLK (blocks), TOV (turnovers), DRB (defensive rebounds), and TRB (total rebounds).

```{r}
#Data Cleaning 
raw_database_24_25 <- read_csv("database_24_25.csv")
head(raw_database_24_25)

```

```{r}
# TEAM-GAME

team_game <- raw_database_24_25 %>%
  group_by(Data, Tm, Opp, Res) %>%  
  summarise(
    FG = sum(FG, na.rm = TRUE),
    FGA = sum(FGA, na.rm = TRUE),
    FG_pct = FG / FGA,
    
    `3P` = sum(`3P`, na.rm = TRUE),
    `3PA` = sum(`3PA`, na.rm = TRUE),
    `3P_pct` = `3P` / `3PA`,
    
    FT = sum(FT, na.rm = TRUE),
    STL = sum(STL, na.rm = TRUE),
    BLK = sum(BLK, na.rm = TRUE),
    
    defensive_pressure = STL + BLK,
    points = sum(PTS, na.rm = TRUE)
  ) %>%
  ungroup()


```

```{r}
head(team_game)
```
```{r}
# TEAM-SEASON

team_season <- team_game %>%
  group_by(Tm) %>%
  summarise(
    wins = sum(Res == "W"),
    losses = sum(Res == "L"),
    avg_FG_pct = mean(FG_pct, na.rm = TRUE),
    avg_3P_pct = mean(`3P_pct`, na.rm = TRUE),
    total_ft = sum(FT, na.rm = TRUE),
    total_stl = sum(STL, na.rm = TRUE),
    total_blk = sum(BLK, na.rm = TRUE),
    avg_def_pressure = mean(defensive_pressure, na.rm = TRUE),
    .groups = "drop"
  )


```


## 3. Exploratory Data Analysis (Tiffany)

```{r}
#Offensive Statistics

```

```{r}
#Defensive Statistics


```


## 4. T-Tests (Shafin)

```{r}

```


## 5. K-Clustering (Derek)

```{r}
team_scaled <- scale(
  team_summary[, c("avg_3P_pct", 
                   "avg_FG_pct", 
                   "total_ft",
                   "total_stl", 
                   "total_blocks")])
#summary(team_scaled)
set.seed(800)
kmeans_result <- kmeans(team_scaled, centers = 3)

team_summary$cluster <- kmeans_result$cluster
```
```{r}
nba_model <- lm(wins ~ avg_3P_pct + avg_FG_pct + total_ft + 
                total_stl + total_blocks, data = team_summary)
summary(nba_model)

plot(team_summary$avg_3P_pct, team_summary$total_stl,
     col = team_summary$cluster,
     pch = 19,
     xlab = "Average 3 Pointer Percentage",
     ylab = "Total Steals",
     main = "3P% and Steals by Team Type")
```

## 6. Logistic Regression (Tiffany)

```{r}
#LOGISTIC REGRESSION MODEL

#Convert results to binary outcome: 1 = Win, 0 = Loss
team_game <- team_game %>%
  mutate(win_binary = ifelse(Res == "W", 1, 0))

# Fit logistic regression predicting win/loss
logit_model <- glm(
  win_binary ~ FG_pct + `3P_pct` + FT + STL + BLK + defensive_pressure + points,
  data = team_game,
  family = binomial
)

# Display model summary
summary(logit_model)
```
## 7. Goodness of Fit Tests (Tiffany)
```{r}
### GOODNESS OF FIT TESTS FOR LOGISTIC REGRESSION ###

# 1. Likelihood Ratio Test (compare model vs null model)
null_model <- glm(win_binary ~ 1, data = team_game, family = binomial)
anova(null_model, logit_model, test = "Chisq")

# 2. Hosmer–Lemeshow Test (requires ResourceSelection package)
if(!require(ResourceSelection)) install.packages("ResourceSelection")
library(ResourceSelection)

hoslem.test(team_game$win_binary, fitted(logit_model), g = 10)

# 3. Pseudo R-squared values (McFadden, Cox & Snell, Nagelkerke)
if(!require(pscl)) install.packages("pscl")
library(pscl)

pR2(logit_model)

# 4. Classification Accuracy and Confusion Matrix
pred_prob <- predict(logit_model, type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

conf_matrix <- table(
  Actual = team_game$win_binary,
  Predicted = pred_class
)
conf_matrix

accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
accuracy

# 5. ROC Curve and AUC
if(!require(pROC)) install.packages("pROC")
library(pROC)

roc_obj <- roc(team_game$win_binary, pred_prob)
plot(roc_obj, main = "ROC Curve — Logistic Regression Model")
auc(roc_obj)

# 6. Brier Score
brier_score <- mean((team_game$win_binary - pred_prob)^2)
brier_score

```
